Imputation of Incomplete Multilevel Data with \pkg{mice}

### 2022-01-19 Manuscript/Manuscript.Rmd
use subset of popularity data as case study

# functions
`%nin%` <- Negate(`%in%`)
icc <- function(formula, data){multilevel::ICC1(aov(formula, data))}
miceadds::source.all(path = "../R")

@@

- `mice::popmis` (simulated data on school kids, with MNAR/MAR mixture);

- `metamisc::impact` (real IPD on traumatic brain injuries, without `NA`s);

@@

```{r pop, echo=FALSE, message=FALSE, warning=FALSE}
# load data 
pop <- readRDS("../Data/popNCR2.RDS") %>% mutate(class = as.factor(class))

@@

`popNCR` is a simulated dataset with pupils clustered in classes, where the number of units $n = 2000$, and the number of clusters $N = 100$, on 7 variables:

@@

```{r pop_pat, echo=FALSE, fig.height=4, fig.width=4, fig.cap = "Missing data pattern in the popularity data", message=FALSE, warning=FALSE}
# missingness
plot_md_pat(pop)
# TODO: add pattern by cluster, text mining example square size
# TODO: try to make it a shepley plot 
```

@@

# compare ICCs before and after imputation
ICCs <- data.frame(
  vars = c("popular", "popteach", "texp"), 
  incomplete = c(icc(popular ~ class, pop), 
               icc(popteach ~ class, pop),
               icc(texp ~ class, pop)), 
  ignored = c(icc(popular ~ class, complete(imp_ignored)), 
              icc(popteach ~ class, complete(imp_ignored)), 
              icc(texp ~ class, complete(imp_ignored)))
  )
ICCs

>>>>

# functions
`%nin%` <- Negate(`%in%`)
icc <- function(formula, data){multilevel::ICC1(aov(formula, data))}
# icc for imputed data as function in mice
miceadds::source.all(path = "../R")

@@

- `mice::popmis` (simulated data on school kids, n = 199 units, N = 10 clusters, with MNAR/MAR mixture);

- `metamisc::impact` (real IPD on traumatic brain injuries, without `NA`s);^[add cross-level interaction?]

@@

```{r pop, echo=FALSE, message=FALSE, warning=FALSE}
# randomly select a few classes
clusters <- sample(1:100, 10)

# load data 
pop <- readRDS("../Data/popNCR2.RDS") %>% 
  filter(class %in% clusters) %>% 
  mutate(class = as.factor(class))

@@

`popNCR2` is a simulated dataset with pupils clustered in classes, where the number of units $n = 2000$, and the number of clusters $N = 100$, on 7 variables:

@@

```{r pop_pat, echo=FALSE, fig.cap = "Missing data pattern in the popularity data", message=FALSE, warning=FALSE}
# missingness
plot_md_pat(pop)
# TODO: add pattern by cluster, text mining example square size
# TODO: try to make it a shepley plot 
```

@@

# compare ICCs before and after imputation
ICCs <- data.frame(
  vars = c("popular", "popteach", "texp"), 
  CCA = c(icc(popular ~ class, pop), 
               icc(popteach ~ class, pop),
               icc(texp ~ class, pop)), 
  ignored = c(icc(popular ~ class, complete(imp_ignored)), 
              icc(popteach ~ class, complete(imp_ignored)), 
              icc(texp ~ class, complete(imp_ignored)))
  )
ICCs

### 2022-01-19 Manuscript/Manuscript.tex
use subset of popularity data as case study

Ignoring the missingness in analyses can be extremely harmful to
inferences. Complete case analysis (i.e., excluding all units with one
or more missing entries) can introduce bias in statistical inference and
lowers statistical power. Instead, the missingness should be
accommodated \emph{before} or \textbf{within} the analysis of scientific
interest. Especially the former is very generic and popular and is
widely known as imputation. Imputing (i.e., filling in) the missing
values separates the missing data problem from the scientific problem:
missing data are replaced by plausible values whereafter the completed
data is analysed as if it were completely observed. The \proglang{R}
package \pkg{mice} has become the de-facto standard for imputation by
chained equations, which iteratively solves the missingness on a
variable-by-variable basis. \pkg{mice} is known to yield valid
inferences under many different missing data circumstances
\citep{buur18}. In this paper, we'll discuss how to use \pkg{mice} in
the context of multilevel data, under varying missing data
mechanisms.\footnote{Discuss missingness mechanisms before this point,
  add references \citet{yuce08} and \citet{hox15}.}

@@

\begin{itemize}
\item
  \texttt{mice::popmis} (simulated data on school kids, with MNAR/MAR
  mixture);
\item
  \texttt{metamisc::impact} (real IPD on traumatic brain injuries,
  without \texttt{NA}s);
\item
  \texttt{GJRM::hiv} (simulated patient data on HIV, without
  \texttt{NA}s)
\end{itemize}

@@

\texttt{popNCR} is a simulated dataset with pupils clustered in classes,
where the number of units \(n = 2000\), and the number of clusters
\(N = 100\), on 7 variables:

@@

The popularity data is created such that there are strong relations
between the incomplete variables and the clustering variable
\texttt{class}. We can express this using the intra-class correlation
(ICC). For \texttt{popular} the ICC is 0.33. For \texttt{popteach} it is
0.34. It would thus be wise to use multilevel modeling.

@@

\begin{CodeOutput}
      vars incomplete   ignored
1  popular  0.3280070 0.3338626
2 popteach  0.3414766 0.3414766
3     texp  1.0000000 1.0000000
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeOutput}
      vars incomplete   ignored predictor
1  popular  0.3280070 0.3338626 0.3871020
2 popteach  0.3414766 0.3414766 0.3414766
3     texp  1.0000000 1.0000000 1.0000000
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> ICCs <- ICCs %>% mutate(
+            norm = c(icc(popular ~ as.factor(class), complete(imp_norm_2l)), 
+                     icc(popteach ~ as.factor(class), complete(imp_norm_2l)), 
+                     icc(texp ~ as.factor(class), complete(imp_norm_2l)))
+            )
R> ICCs
\end{CodeInput}
\begin{CodeOutput}
      vars incomplete   ignored predictor      norm
1  popular  0.3280070 0.3338626 0.3871020 0.3612830
2 popteach  0.3414766 0.3414766 0.3414766 0.3414766
3     texp  1.0000000 1.0000000 1.0000000 1.0000000
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> ICCs <- ICCs %>% mutate(
+            pan = c(icc(popular ~ as.factor(class), complete(imp_pan_2l)), 
+                     icc(popteach ~ as.factor(class), complete(imp_pan_2l)), 
+                     icc(texp ~ as.factor(class), complete(imp_pan_2l)))
+            )
R> ICCs
\end{CodeInput}
\begin{CodeOutput}
      vars incomplete   ignored predictor      norm       pan
1  popular  0.3280070 0.3338626 0.3871020 0.3612830 0.3745108
2 popteach  0.3414766 0.3414766 0.3414766 0.3414766 0.3414766
3     texp  1.0000000 1.0000000 1.0000000 1.0000000 1.0000000
\end{CodeOutput}
\end{CodeChunk}

>>>>

Ignoring the missingness in analyses can be extremely harmful to
inferences. Complete case analysis (i.e., excluding all units with one
or more missing entries) can introduce bias in statistical inference and
lowers statistical power. Instead, the missingness should be
accommodated \emph{before} or \emph{within} the analysis of scientific
interest. Especially the former is very generic and popular and is
widely known as imputation. Imputing (i.e., filling in) the missing
values separates the missing data problem from the scientific problem:
missing data are replaced by plausible values whereafter the completed
data is analysed as if it were completely observed. The \proglang{R}
package \pkg{mice} has become the de-facto standard for imputation by
chained equations, which iteratively solves the missingness on a
variable-by-variable basis. \pkg{mice} is known to yield valid
inferences under many different missing data circumstances
\citep{buur18}. In this paper, we'll discuss how to use \pkg{mice} in
the context of multilevel data, under varying missing data
mechanisms.\footnote{Discuss missingness mechanisms before this point,
  add references \citet{yuce08} and \citet{hox15}.}

@@

\begin{itemize}
\item
  \texttt{mice::popmis} (simulated data on school kids, n = 199 units, N
  = 10 clusters, with MNAR/MAR mixture);
\item
  \texttt{metamisc::impact} (real IPD on traumatic brain injuries,
  without \texttt{NA}s);\footnote{add cross-level interaction?}
\item
  \texttt{GJRM::hiv} (simulated patient data on HIV, without
  \texttt{NA}s)
\end{itemize}

@@

\texttt{popNCR2} is a simulated dataset with pupils clustered in
classes, where the number of units \(n = 2000\), and the number of
clusters \(N = 100\), on 7 variables:

@@

The popularity data is created such that there are strong relations
between the incomplete variables and the clustering variable
\texttt{class}. We can express this using the intra-class correlation
(ICC). For \texttt{popular} the ICC is 0.4. For \texttt{popteach} it is
0.36. It would thus be wise to use multilevel modeling.

@@

\begin{CodeOutput}
      vars       CCA   ignored
1  popular 0.3989463 0.4235643
2 popteach 0.3605902 0.3605902
3     texp 1.0000000 1.0000000
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeOutput}
      vars       CCA   ignored predictor
1  popular 0.3989463 0.4235643 0.3904372
2 popteach 0.3605902 0.3605902 0.3605902
3     texp 1.0000000 1.0000000 1.0000000
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> ICCs <- ICCs %>% mutate(
+            norm = c(icc(popular ~ as.factor(class), complete(imp_norm_2l)), 
+                     icc(popteach ~ as.factor(class), complete(imp_norm_2l)), 
+                     icc(texp ~ as.factor(class), complete(imp_norm_2l)))
+            )
R> ICCs
\end{CodeInput}
\begin{CodeOutput}
      vars       CCA   ignored predictor      norm
1  popular 0.3989463 0.4235643 0.3904372 0.4165006
2 popteach 0.3605902 0.3605902 0.3605902 0.3605902
3     texp 1.0000000 1.0000000 1.0000000 1.0000000
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> ICCs <- ICCs %>% mutate(
+            pan = c(icc(popular ~ as.factor(class), complete(imp_pan_2l)), 
+                     icc(popteach ~ as.factor(class), complete(imp_pan_2l)), 
+                     icc(texp ~ as.factor(class), complete(imp_pan_2l)))
+            )
R> ICCs
\end{CodeInput}
\begin{CodeOutput}
      vars       CCA   ignored predictor      norm       pan
1  popular 0.3989463 0.4235643 0.3904372 0.4165006 0.3827595
2 popteach 0.3605902 0.3605902 0.3605902 0.3605902 0.3605902
3     texp 1.0000000 1.0000000 1.0000000 1.0000000 1.0000000
\end{CodeOutput}
\end{CodeChunk}
